---
title: "CPSC540_Homework1"
output:
  html_document:
    df_print: paged
date: "2024-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(dagitty)
library(ggdag)
```

```{r}
data <- read.csv("Documents/CADS FALL 24 (FINAL SEM)/CPSC540/data.csv", sep = ";", header = TRUE, stringsAsFactors = FALSE)
str(data)
```

```{r}
summary(data)
```
Data Cleaning:
```{r}
# check for missing values
missing_values <- sapply(data, function(x) sum(is.na(x)))
missing_values
```

Rename columns with special characters
```{r}
# rename columns to remove apostraphes
data <- data %>%
  rename(
    Mothers_qualification = `Mother.s.qualification`,
    Fathers_qualification = `Father.s.qualification`,
    Mothers_occupation = `Mother.s.occupation`,
    Fathers_occupation = `Father.s.occupation`
  )
```

```{r}
# standardize column names by replacing non-standard characters with underscores
colnames(data) <- make.names(colnames(data))
```

```{r}
# conert columns to factors and numeric
data <- data %>%
  mutate(
    Daytime.evening.attendance = as.factor(Daytime.evening.attendance.),
    Daytime.evening.attendance. = recode(Daytime.evening.attendance., 
                                         `1` = "Daytime", 
                                         `0` = "Evening"),
    Age.at.enrollment = as.numeric(Age.at.enrollment),
    Admission.grade = as.numeric(Admission.grade),
    Scholarship.holder = as.factor(Scholarship.holder),
    Tuition.fees.up.to.date = as.factor(Tuition.fees.up.to.date),
    Curricular.units.1st.sem..grade. = as.numeric(Curricular.units.1st.sem..grade.),
    Target = as.factor(Target)
  )
```

Exploratory Data Analysis


```{r}
# Define the DAG with all relevant variables
dag <- dagitty("dag {
  Attendance -> Dropout
  Age -> Attendance
  Age -> Dropout
  Admission -> Attendance
  Admission -> Dropout
  Scholarship -> Attendance
  Scholarship -> Dropout
  Tuition_UpToDate -> Attendance
  Tuition_UpToDate -> Dropout
  First_Sem_Grade -> Attendance
  First_Sem_Grade -> Dropout
}")

ggdag(dag, text = TRUE) +
  theme_minimal() +
  labs(title = "DAG for Attendance Type and Dropout Relationship") +
  theme(legend.position = "none")
```


Plot 1: Distribution of Attendance Type
```{r}
ggplot(data, aes(x = Daytime.evening.attendance.)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of Daytime and Evening Students", x = "Attendance Type", y = "Count") +
  scale_x_discrete(labels = c("Daytime", "Evening")) +
  theme_minimal()
```

```{r}
# recode Target variable as binary
# Dropout = 1 and others = 0
data <- data %>%
  mutate(Target = ifelse(Target == "Dropout", 1, 0))
```

```{r}
# calculate dropout rate by attendance type
dropout_rate <- data_filtered %>%
  group_by(Daytime.evening.attendance.) %>%
  summarize(dropout_rate = mean(Target), .groups = 'drop')

print(dropout_rate)
```
Plot 2: Dropout Rate by Attendance Type
```{r}
# Plot dropout rates by attendance type
ggplot(dropout_rate, aes(x = Daytime.evening.attendance., y = dropout_rate)) +
  geom_col(fill = "coral") +
  labs(title = "Dropout Rate by Attendance Type", x = "Attendance Type", y = "Dropout Rate") +
  theme_minimal()
```
Plot 3: Socioeconomic Status and Dropout Relationship
```{r}
# Age at Enrollment plot
ggplot(data, aes(x = Age.at.enrollment)) +
  geom_histogram(bins = 15, fill = "lightgreen") +
  labs(title = "Age Distribution at Enrollment", x = "Age", y = "Count") +
  theme_minimal()

# Admission Grade plot
ggplot(data, aes(x = Admission.grade)) +
  geom_histogram(bins = 20, fill = "lightblue") +
  labs(title = "Admission Grade Distribution", x = "Admission Grade", y = "Count") +
  theme_minimal()
```
Baseline Logistic Regression

```{r}
# Baseline logistic regression model
baseline_model <- glm(Target ~ Daytime.evening.attendance. + Age.at.enrollment + Admission.grade +
                        Scholarship.holder + Tuition.fees.up.to.date + Curricular.units.1st.sem..grade., 
                      data = data_filtered, 
                      family = binomial)
```

```{r}
summary(baseline_model)
# print odds ratio
exp(coef(baseline_model))
```

Implementing Inverse Probability of Treatment Weighting (IPTW)
```{r}
# Recode Daytime.evening.attendance. as binary (0 = Daytime, 1 = Evening)
data_filtered <- data_filtered %>%
  mutate(Daytime.evening.attendance. = ifelse(Daytime.evening.attendance. == "Evening", 1, 0))

# Calculate Propensity Scores
propensity_model <- glm(Daytime.evening.attendance. ~ Age.at.enrollment + Admission.grade +
                         Scholarship.holder + Tuition.fees.up.to.date + Curricular.units.1st.sem..grade., 
                        data = data_filtered, family = binomial)

# Calculate weights based on propensity scores
data_filtered$weights <- ifelse(data_filtered$Daytime.evening.attendance. == 1,
                                1 / predict(propensity_model, type = "response"),
                                1 / (1 - predict(propensity_model, type = "response")))

# Fit Weighted Logistic Regression Model
iptw_model <- glm(Target ~ Daytime.evening.attendance., data = data_filtered, 
                  family = binomial, weights = weights)
summary(iptw_model)

# Odds Ratios
exp(coef(iptw_model))
```

Visualizations:
1. Propensity Score Distribution by Attendance Type
```{r}
library(ggplot2)

# calculate propensity scores
data_filtered$propensity_score <- predict(propensity_model, type = "response")

# plot density of propensity scores by attendance type
ggplot(data_filtered, aes(x = propensity_score, fill = as.factor(Daytime.evening.attendance.))) +
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Propensity Scores by Attendance Type", 
       x = "Propensity Score", y = "Density", fill = "Attendance Type") +
  theme_minimal()
```
2. Weighted Dropout Rates by Attendance Type

```{r}
# calculate weighted dropout rates by attendance type
weighted_dropout_rate <- data_filtered %>%
  group_by(Daytime.evening.attendance.) %>%
  summarize(weighted_dropout = weighted.mean(Target, weights))

# plot weighted dropout rates
ggplot(weighted_dropout_rate, aes(x = as.factor(Daytime.evening.attendance.), y = weighted_dropout)) +
  geom_col(fill = "coral") +
  labs(title = "Weighted Dropout Rates by Attendance Type", 
       x = "Attendance Type", y = "Weighted Dropout Rate") +
  theme_minimal()

```